A Tutorial on Tailored Simulation-Based Sample Size Planning for Experimental Designs with Generalized Linear Mixed Models

### 2024-11-05 _manuscript/_tex/manuscript.tex
minor updates based on proof-reading; include citation notice; reformat precision results figure

\par{ This is version 3.5 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest concerning the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant No.~98525).  Florian
Pargent and Timo Koch contributed equally to this
manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
acquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

@@

First, requirements of real-world studies are often more complex than
the simplified designs assumed by many user-friendly software packages
for sample size planning. One frequent issue in applied data analysis is
missing data (\citeproc{ref-littleStatisticalAnalysisMissing2014}{Little
\& Rubin, 2014}). Missings can be completely random (e.g., an electronic
measurement device randomly failed for technical reasons), depend on
some attributes also measured in the dataset (e.g., older subjects
refuse answering a question on income), or be caused by the measured
attribute itself (e.g., wealthy people refuse to report their income).
Moreover, many experimental designs contain conditions in which values
of the predictor variables are missing by design. This can make data
analysis more complicated because predictors have to be coded in
specific ways that prevent the estimated GLMM from becoming
unidentified. Whether missing data has an effect on the sample size
planning depends on our theoretical assumptions on how the missingness
is caused. However, it is often challenging to decide whether missing
data can be safely ignored in the data analysis and sample size planning
process based on a merely theoretical approach
(\citeproc{ref-gomilaMissingDataExperiments2022}{Gomila \& Clark,
2022}). Tailored simulation-based approaches offer the possibility to
include the assumed process of how data become missing in the data
simulation, thereby determining the required sample size based on
simulated datasets that contain missing values (for example, see
\citeproc{ref-lanePowerStrugglesEstimating2018}{Lane \& Hennes, 2018}).
As a byproduct, the simulated datasets can also be used to test whether
the intended data analysis provides the expected (unbiased) results,
despite the missing data. Although GLMMs can handle a large variety of
outcome variables, researchers are becoming increasingly aware that many
datasets might profit from even more sophisticated models. Common
examples are zero-inflated outcomes, censoring, and nonlinear predictor
effects that can be modeled with the R packages \emph{glmmTMB}
(\citeproc{ref-brooksGlmmTMBBalancesSpeed2017}{Brooks et al., 2017}) or
\emph{brms}
(\citeproc{ref-burknerAdvancedBayesianMultilevel2018}{Bürkner, 2018}).
Tailored simulation-based approaches do not share the same limitations
as the existing software solutions for power analysis that focus
exclusively on GLMMs. As long as there is a software package available
to estimate the model of interest, it is always possible to perform
tailored simulation-based sample size planning.

Second, the most common hypotheses tested in psychological research are
of the type \(H_0: \beta = 0\), where \(\beta\) is a slope or intercept
of a regression model. However, many research questions in psychology
actually require testing more complex statistical hypotheses. In the new
era of preregistration and registered reports
(\citeproc{ref-chambersPresentFutureRegistered2022}{Chambers \&
Tzavella, 2022}), most research questions should be tested with directed
hypotheses because good theories at least postulate whether some
psychological effect of interest is positive or negative. Even better
theories should be able to specify the smallest effect sizes of interest
(SESOI) that must be exceeded if the effect has any practical relevance
(\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens,
Scheel, et al., 2018}). This might require a test such as
\(H_0: \beta \leq 0.1\). More elaborate research questions often require
testing hypotheses that consist of a combination of model parameters,
for example, testing simple slopes
(\citeproc{ref-preacherComputationalToolsProbing2006}{Preacher et al.,
2006}) with a hypothesis such as \(H_0: \beta_0 + \beta_1 \leq 0\). If
the research question consists only of a single hypothesis of this sort,
it might be possible to reduce the hypothesis to a single regression
coefficient by clever coding and/or centering of predictor variables.
However, research questions often consist of combined hypotheses that
consist of more than one separate statistical hypothesis (for a tutorial
on contrast analysis in GLMMs, see
\citeproc{ref-schadHowCapitalizePriori2020}{Schad et al., 2020}). For
example, a combined null hypothesis \(H_0\) might consist of two single
null hypotheses \(H_{01}: \beta_1 \leq 0\) and
\(H_{02}: \beta_0 + \beta_1 \leq 0\). For some research questions, the
combined null hypothesis \(H_0\) would be rejected if both \(H_{01}\)
\emph{AND} \(H_{02}\) are rejected. For other research questions, the
combined null hypothesis \(H_0\) would be rejected if \(H_{01}\)
\emph{OR} \(H_{02}\) \emph{OR} both are rejected. If the global
hypothesis \(H_0\) is combined with \emph{OR}, the p-values of the
single hypotheses must be corrected for multiple testing to avoid
\(\alpha\)-inflation for the global hypothesis
(\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). However, if the global hypothesis \(H_0\) is
combined with \emph{AND}, a correction for multiple testing is not
necessary but rather a mistake that unnecessarily reduces the power of
the global HT. None of the software packages for sample size planning in
Table 1 can handle combined hypotheses and only some can handle directed
hypotheses. In contrast, our case study will demonstrate how we can test
directed combined hypotheses with tailored simulation-based sample size
planning.

@@

In such a setting, we model this probability as \[
P_{si} = \text{inverse\_logit}(\eta_{si})
\] with the inverse-logit link
\(g^{-1}(\eta_{si}) = inverse\_logit(\eta_{si}) = \frac{exp(\eta_{si})}{1 + exp(\eta_{si})}\)
or equivalently \[
\text{logit}(P_{si}) = \eta_{si}
\] with the logit link
\(g(P_{si}) = \text{logit}(P_{si}) = \text{ln} (\frac{P_{si}}{1 - P_{si}})\).

@@

In total, we assume the model \[
\begin{aligned}
\text{logit}[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
\] or equivalently \[
\begin{aligned}
P_{si} = \text{inverse\_logit}[&(\beta_0 + u_{0s} + u_{0i}) + \\
&\beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
\] with model parameters \(\beta_0\), \(\beta_e\), \(\beta_a\),
\(\beta_c\), \(\beta_{ea}\), \(\beta_{ec}\), \(\sigma_S\), and
\(\sigma_I\).

@@

\[
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0}) = \\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s + \\
&\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
\] In fact, we would only need the full equation if the subject is an
expert and correct advice is presented. In all other experimental
conditions, some terms drop from the equation because they are
multiplied by \(0\). For example, the probability that a student with
average ability solves an item with average difficulty when no advice is
presented only requires the intercept: \[
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0) = \\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
\] We can revert this perspective by choosing plausible probability
values based on domain knowledge and deriving the parameter values
implied by these probabilities for each experimental condition.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1646}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3291}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5063}}@{}}
\caption{Table 2. Assumed probabilities that an average subject solves
an average item in each experimental condition.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Experimental condition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(P(Y=1 \mid \mathbf{X=x}, \mathbf{U} = \mathbf{0})\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implied equation
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Experimental condition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(P(Y=1 \mid \mathbf{X=x}, \mathbf{U} = \mathbf{0})\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implied equation
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
no advice, student & 0.70 & \(logit(0.70) = \beta_0\) \\
no advice, expert & 0.90 & \(logit(0.90) = \beta_0 + \beta_e\) \\
incorrect advice, student & 0.40 &
\(logit(0.40) = \beta_0 + \beta_a\) \\
incorrect advice, expert & 0.85 &
\(logit(0.85) = \beta_0 + \beta_e + \beta_{a} + \beta_{ea}\) \\
correct advice, student & 0.90 &
\(logit(0.90) = \beta_0 + \beta_a + \beta_c\) \\
correct advice, expert & 0.95 &
\(logit(0.95) = \beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}\) \\
\end{longtable}

@@

It is always possible to double-check these computations by transforming
the parameter values back to probabilities, e.g.~ \[
\begin{aligned}
P(Y=1|expert = 1, advice\_present = 1, advice\_correct = 1, u_{0s} = 0, u_{0i} = 0) = \\
= inverse\_logit[\beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}]
\end{aligned}
\]

@@

For example, we determined the \(\beta\) parameters in a way that
corresponds to a desired conditional probability of \(0.95\), that an
expert with average ability solves an item with average difficulty when
presented with correct advice (the conditional perspective). However,
even if the model assumptions were true, we would not observe that 95\%
of experts responding to items presented with correct advice from a big
sample of subjects drawn from their natural distribution of ability and
items drawn from their natural distribution of difficulty (the marginal
perspective). How much the two probabilities differ depends on the
standard deviations of the random intercepts (the two probabilities are
only equal if both standard deviations are be zero). We want to use the
model-implied observed proportion of correct diagnoses in each
experimental condition as an insightful descriptive statistic to
determine plausible values for the random effect standard deviations. We
will simulate a large dataset (for which the observed values of the
descriptive statistic will be close to their model-implied true values)
and simply compute the relative frequency of correct diagnoses for each
experimental condition.

>>>>

\par{ This is version 4 of our preprint published at
\url{https://doi.org/10.31234/osf.io/rpjem}. All materials (reproducible
manuscript, including R code and all simulation results) are available
in the project's repository on the Open Science Framework (OSF) at
\url{https://osf.io/dhwf4/}. A Quarto Manuscripts website is hosted at
\url{https://Timo-Ko.github.io/glmm_simulation_tutorial/}.  The authors
declare that there were no conflicts of interest concerning the
authorship or the publication of this article. This research was funded
by a grant from the Volkswagen Foundation (Grant No.~98525).  Florian
Pargent and Timo Koch contributed equally to this
manuscript. Author roles were classified using the Contributor Role Taxonomy (CRediT; https://credit.niso.org/) as follows: Florian
Pargent:   formal
analysis, methodology, visualization, writing, editing; Timo K.
Koch:   conceptualization, formal
analysis, writing, editing; Anne-Kathrin Kleine:   formal
analysis, editing; Eva Lermer:   conceptualization, funding
acquisition, editing; Susanne
Gaube:   conceptualization, supervision, editing}
\par{Correspondence concerning this article should be addressed to Timo
K. Koch, Institute of Behavioral Science \& Technology, University of
St.~Gallen, Torstrasse
25, St.~Gallen 9000, Switzerland, Email: timo.koch@unisg.ch}
}

@@

\usepackage{fontspec}
\usepackage{multirow}
\usepackage{multicol}
\usepackage{colortbl}
\usepackage{hhline}
\newlength\Oldarrayrulewidth
\newlength\Oldtabcolsep
\usepackage{longtable}
\usepackage{array}
\usepackage{hyperref}
\usepackage{float}
\usepackage{wrapfig}
\usepackage{tabularray}
\usepackage[normalem]{ulem}
\usepackage{graphicx}
\UseTblrLibrary{booktabs}
\UseTblrLibrary{siunitx}
\NewTableCommand{\tinytableDefineColor}[3]{\definecolor{#1}{#2}{#3}}
\newcommand{\tinytableTabularrayUnderline}[1]{\underline{#1}}
\newcommand{\tinytableTabularrayStrikeout}[1]{\sout{#1}}
\makeatletter
\@ifpackageloaded{tcolorbox}{}{\usepackage[skins,breakable]{tcolorbox}}
\@ifpackageloaded{fontawesome5}{}{\usepackage{fontawesome5}}
\definecolor{quarto-callout-color}{HTML}{909090}
\definecolor{quarto-callout-note-color}{HTML}{0758E5}
\definecolor{quarto-callout-important-color}{HTML}{CC1914}
\definecolor{quarto-callout-warning-color}{HTML}{EB9113}
\definecolor{quarto-callout-tip-color}{HTML}{00A047}
\definecolor{quarto-callout-caution-color}{HTML}{FC5300}
\definecolor{quarto-callout-color-frame}{HTML}{acacac}
\definecolor{quarto-callout-note-color-frame}{HTML}{4582ec}
\definecolor{quarto-callout-important-color-frame}{HTML}{d9534f}
\definecolor{quarto-callout-warning-color-frame}{HTML}{f0ad4e}
\definecolor{quarto-callout-tip-color-frame}{HTML}{02b875}
\definecolor{quarto-callout-caution-color-frame}{HTML}{fd7e14}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother

@@

\begin{tcolorbox}[enhanced jigsaw, leftrule=.75mm, arc=.35mm, colframe=quarto-callout-important-color-frame, toprule=.15mm, colback=white, title=\textcolor{quarto-callout-important-color}{\faExclamation}\hspace{0.5em}{Important}, bottomtitle=1mm, opacitybacktitle=0.6, colbacktitle=quarto-callout-important-color!10!white, left=2mm, opacityback=0, rightrule=.15mm, breakable, coltitle=black, toptitle=1mm, titlerule=0mm, bottomrule=.15mm]

This manuscript has been accepted by the journal \emph{Advances in
Methods and Practices in Psychological Science (AMPPS)}. Please cite the
official version of record as:

Pargent, F., K. Koch, T., Kleine, A.-K., Lermer, E., \& Gaube, S.
(2024). A tutorial on tailored simulation-based sample-size planning for
experimental designs with generalized linear mixed models.
\emph{Advances in Methods and Practices in Psychological Science}, 7(4).
\url{https://doi.org/10.1177/25152459241287132}

\end{tcolorbox}

@@

First, requirements of real-world studies are often more complex than
the simplified designs assumed by many user-friendly software packages
for sample size planning. One frequent issue in applied data analysis is
missing data (\citeproc{ref-littleStatisticalAnalysisMissing2014}{Little
\& Rubin, 2014}). Missings can be completely random (e.g., an electronic
measurement device randomly failed for technical reasons), depend on
some attributes also measured in the dataset (e.g., older subjects
refuse answering a question on income), or be caused by the measured
attribute itself (e.g., wealthy people refuse to report their income).
Moreover, many experimental designs contain conditions in which values
of the predictor variables are missing by design. This can make data
analysis more complicated because predictors have to be coded in
specific ways that prevent the estimated GLMM from becoming
unidentified. Whether missing data has an effect on the sample size
planning depends on our theoretical assumptions on how the missingness
is caused. However, it is often challenging to decide whether missing
data can be safely ignored in the data analysis and sample size planning
process based on a merely theoretical approach
(\citeproc{ref-gomilaMissingDataExperiments2022}{Gomila \& Clark,
2022}). Tailored simulation-based approaches offer the possibility to
include the assumed process of how data become missing in the data
simulation, thereby determining the required sample size based on
simulated datasets that contain missing values (for example, see
\citeproc{ref-lanePowerStrugglesEstimating2018}{Lane \& Hennes, 2018}).
As a by-product, the simulated datasets can also be used to test whether
the intended data analysis provides the expected (unbiased) results,
despite the missing data. Although GLMMs can handle a large variety of
outcome variables, researchers are becoming increasingly aware that many
datasets might profit from even more sophisticated models. Common
examples are zero-inflated outcomes, censoring, and nonlinear predictor
effects that can be modeled with the R packages \emph{glmmTMB}
(\citeproc{ref-brooksGlmmTMBBalancesSpeed2017}{Brooks et al., 2017}) or
\emph{brms}
(\citeproc{ref-burknerAdvancedBayesianMultilevel2018}{Bürkner, 2018}).
Tailored simulation-based approaches do not share the same limitations
as the existing software solutions for power analysis that focus
exclusively on GLMMs. As long as there is a software package available
to estimate the model of interest, it is always possible to perform
tailored simulation-based sample size planning.

Second, the most common hypotheses tested in psychological research are
of the type \(H_0: \beta = 0\), where \(\beta\) is a slope or intercept
of a regression model. However, many research questions in psychology
actually require testing more complex statistical hypotheses. In the new
era of preregistration and registered reports
(\citeproc{ref-chambersPresentFutureRegistered2022}{Chambers \&
Tzavella, 2022}), most research questions should be tested with directed
hypotheses because good theories at least postulate whether some
psychological effect of interest is positive or negative. Even better
theories should be able to specify the smallest effect sizes of interest
(SESOI) that must be exceeded if the effect has any practical relevance
(\citeproc{ref-lakensEquivalenceTestingPsychological2018}{Lakens,
Scheel, et al., 2018}). This might require a test such as
\(H_0: \beta \leq 0.1\). More elaborate research questions often require
testing hypotheses that consist of a combination of model parameters,
for example, testing simple slopes
(\citeproc{ref-preacherComputationalToolsProbing2006}{Preacher et al.,
2006}) with a hypothesis such as \(H_0: \beta_0 + \beta_1 \leq 0\). If
the research question consists only of a single hypothesis of this sort,
it might be possible to reduce the hypothesis to a single regression
coefficient by clever coding and/or centering of predictor variables.
However, research questions often consist of combined hypotheses that
consist of more than one separate statistical hypothesis (for a tutorial
on contrast analysis in GLMMs, see
\citeproc{ref-schadHowCapitalizePriori2020}{Schad et al., 2020}). For
example, a combined null hypothesis \(H_0\) might consist of two single
null hypotheses \(H_{01}: \beta_1 \leq 0\) and
\(H_{02}: \beta_0 + \beta_1 \leq 0\). For some research questions, the
combined null hypothesis \(H_0\) would be rejected if both \(H_{01}\)
\emph{AND} \(H_{02}\) are rejected. For other research questions, the
combined null hypothesis \(H_0\) would be rejected if \(H_{01}\)
\emph{OR} \(H_{02}\) \emph{OR} both are rejected. If the global
hypothesis \(H_0\) is combined with a logical \emph{OR}, the p-values of
the single hypotheses must be corrected for multiple testing to avoid
\(\alpha\)-inflation for the global hypothesis
(\citeproc{ref-dmitrienkoTraditionalMultiplicityAdjustment2013}{Dmitrienko
\& D'Agostino, 2013}). However, if the global hypothesis \(H_0\) is
combined with a logical \emph{AND}, a correction for multiple testing is
not necessary but rather a mistake that unnecessarily reduces the power
of the global HT. None of the software packages for sample size planning
in Table 1 can handle combined hypotheses and only some can handle
directed hypotheses. In contrast, our case study will demonstrate how we
can test directed combined hypotheses with tailored simulation-based
sample size planning.

@@

In such a setting, we model this probability as \[
P_{si} = \text{inverse\_logit}(\eta_{si})
\] with the inverse-logit link
\(g^{-1}(\eta_{si}) = \text{inverse\_logit}(\eta_{si}) = \frac{\text{exp}(\eta_{si})}{1 + \text{exp}(\eta_{si})}\)
or equivalently \[
\text{logit}(P_{si}) = \eta_{si}
\] with the logit link
\(g(P_{si}) = \text{logit}(P_{si}) = \text{ln} (\frac{P_{si}}{1 - P_{si}})\).

@@

In total, we assume the model \[
\begin{aligned}
\text{logit}[P_{si}] =\ (&\beta_0 + u_{0s} + u_{0i})\\
&+ \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s\\
&+\beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}
\end{aligned}
\] or equivalently \[
\begin{aligned}
P_{si} = \text{inverse\_logit}[&(\beta_0 + u_{0s} + u_{0i})\\
&+ \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} + \beta_e \cdot expert_s\\
&+ \beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
\] with model parameters \(\beta_0\), \(\beta_e\), \(\beta_a\),
\(\beta_c\), \(\beta_{ea}\), \(\beta_{ec}\), \(\sigma_S\), and
\(\sigma_I\).

@@

\[
\begin{aligned}
P(Y=1|\mathbf{X=x}, \mathbf{U} = \mathbf{0})\\
= \text{inverse\_logit}[&\beta_0 + \beta_a \cdot advice\_present_{si} + \beta_c \cdot advice\_correct_{si} +  \beta_e \cdot expert_s\\
&+ \beta_{ea} \cdot expert_{s} \cdot advice\_present_{si} + \beta_{ec} \cdot expert_{s} \cdot advice\_correct_{si}]
\end{aligned}
\] In fact, we would only need the full equation if the subject is an
expert and correct advice is presented. In all other experimental
conditions, some terms drop from the equation because they are
multiplied by \(0\). For example, the probability that a student with
average ability solves an item with average difficulty when no advice is
presented only requires the intercept: \[
\begin{aligned}
P(Y=1| advice\_present = 0, advice\_correct = 0, expert = 0, u_{0s} = 0, u_{0i} = 0)\\
= \text{inverse\_logit}[\beta_0]
\end{aligned}
\] We can revert this perspective by choosing plausible probability
values based on domain knowledge and deriving the parameter values
implied by these probabilities for each experimental condition.

@@

\begin{longtable}[]{@{}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.1576}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.3152}}
  >{\raggedright\arraybackslash}p{(\columnwidth - 4\tabcolsep) * \real{0.5273}}@{}}
\caption{Table 2. Assumed probabilities that an average subject solves
an average item in each experimental condition.}\tabularnewline
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Experimental condition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(P(Y=1 \mid \mathbf{X=x}, \mathbf{U} = \mathbf{0})\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implied equation
\end{minipage} \\
\midrule\noalign{}
\endfirsthead
\toprule\noalign{}
\begin{minipage}[b]{\linewidth}\raggedright
Experimental condition
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
\(P(Y=1 \mid \mathbf{X=x}, \mathbf{U} = \mathbf{0})\)
\end{minipage} & \begin{minipage}[b]{\linewidth}\raggedright
Implied equation
\end{minipage} \\
\midrule\noalign{}
\endhead
\bottomrule\noalign{}
\endlastfoot
no advice, student & 0.70 & \(\text{logit}(0.70) = \beta_0\) \\
no advice, expert & 0.90 & \(\text{logit}(0.90) = \beta_0 + \beta_e\) \\
incorrect advice, student & 0.40 &
\(\text{logit}(0.40) = \beta_0 + \beta_a\) \\
incorrect advice, expert & 0.85 &
\(\text{logit}(0.85) = \beta_0 + \beta_e + \beta_{a} + \beta_{ea}\) \\
correct advice, student & 0.90 &
\(\text{logit}(0.90) = \beta_0 + \beta_a + \beta_c\) \\
correct advice, expert & 0.95 &
\(\text{logit}(0.95) = \beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}\) \\
\end{longtable}

@@

It is always possible to double-check these computations by transforming
the parameter values back to probabilities, e.g.~ \[
\begin{aligned}
P(Y=1|expert = 1, advice\_present = 1, advice\_correct = 1, u_{0s} = 0, u_{0i} = 0)\\
= \text{inverse\_logit}[\beta_0 + \beta_e + \beta_a + \beta_c + \beta_{ea} + \beta_{ec}]
\end{aligned}
\]

@@

For example, we determined the \(\beta\) parameters in a way that
corresponds to a desired conditional probability of \(0.95\), that an
expert with average ability solves an item with average difficulty when
presented with correct advice (the conditional perspective). However,
even if the model assumptions were true, the relative frequency of
experts that solve items presented with correct advice would not be
\(0.95\) in a large sample of subjects drawn from their natural
distribution of ability and items drawn from their natural distribution
of difficulty (the marginal perspective). How much the two probabilities
differ depends on the standard deviations of the random intercepts (the
two probabilities are only equal if both standard deviations are be
zero). We want to use the model-implied observed proportion of correct
diagnoses in each experimental condition as an insightful descriptive
statistic to determine plausible values for the random effect standard
deviations. We will simulate a large dataset (for which the observed
values of the descriptive statistic will be close to their model-implied
true values) and simply compute the relative frequency of correct
diagnoses for each experimental condition.
